# Chapter 6

This is the last chapter and the assignment of this course:
For this analysis section we will be working with the BPRS and RATS data but swap the methods used for each previously in the exercise set 6.

Part 1: RATS DATASET
We are starting with the RATS dataset.This is from a nutrition study conducted in three groups of rats (Crowder and Hand, 1990). The rats were grouped into three and were given different diets and their weights were measured over a 9 week period.


```{r}
RATS <- read.table("https://raw.githubusercontent.com/KimmoVehkalahti/MABS/master/Examples/data/rats.txt", header = TRUE, sep = '\t')

dim(RATS) # 16 observations and 13 variables.
RATS
# Have to turn the categorical data into factors:
RATS$ID = factor(RATS$ID)
RATS$Group = factor(RATS$Group)

# Turn RATS into RATSL by turning it into a long data set
library(dplyr)
library(tidyr)

# Convert data to long form
RATSL <- pivot_longer(RATS, cols = -c(ID, Group), 
                      names_to = "WD",
                      values_to = "Weight") %>% 
         mutate(Time = as.integer(substr(WD,3,4))) %>%
         arrange(Time)

# Have a look at the data 
dim(RATSL) # 176 observations and 5 variables. Successfully turned into longer dataset.
RATSL

```

```{r}
#Access the packages
library(ggplot2)

summary(RATSL)

# Draw the plot
ggplot(RATSL, aes(x = Time, y = Weight, linetype = ID)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ Group, labeller = label_both) +
  theme(legend.position = "none") + 
  scale_y_continuous(limits = c(min(RATSL$Weight), max(RATSL$Weight)))

# Trying also a boxplot:
ggplot(RATSL, aes(x = Group, y = Weight, fill = Group)) +
  geom_boxplot() +
  facet_wrap(~WD, scales = "free") +
  scale_fill_discrete(name = "Group") +
  theme_minimal()

```

Here we can see that the group 1 treat ment is doing the worst in terms of weight gain and group 3 is kind of the best as it is consistently doing better with weight increase.

```{r}
library(dplyr)
library(tidyr)

# Standardize data
RATSS <- RATSL %>%
  group_by(Time) %>%
  mutate(stdwght = (Weight - mean(Weight))/sd(Weight)) %>%
  ungroup()

# Glimpse the data
glimpse(RATSS)

# Calculate mean and standard error for Group and Time
RATSS_s <- RATSL %>%
  group_by(Group, Time) %>%
    summarise(mean = mean(Weight), n = n(), se = sd(Weight)/sqrt(n)) %>%
  ungroup()

# Glimpse the data
glimpse(RATSS_s)

# Plot the mean profiles with error bars
library(ggplot2)

# Plotting the mean profiles
ggplot(RATSS_s, aes(x = Time, y = mean, linetype = Group, shape = Group)) +
  geom_line() +
  scale_linetype_manual(values = c(1,2,3)) +
  geom_point(size=3) +
  scale_shape_manual(values = c(1,2,3)) +
  geom_errorbar(aes(ymin=mean-se, ymax=mean+se, linetype="1"), width=0.3) +
  theme(legend.position = c(0.8,0.5)) +
  scale_y_continuous(name = "mean(weight) +/- se(weight)")

```

```{r}
# We have a look for outliers
library(dplyr)
library(ggplot2)

# Create a summary data by Group and WD with mean as the summary variable

RATSLs <- RATSL %>%
  filter(Time > 1) %>%
  group_by(Group, ID) %>%
  summarise(mean=mean(Weight) ) %>%
  ungroup()

# Glimpse the data
glimpse(RATSLs)

# Drawing a boxplot for checking outliers
ggplot(RATSLs, aes(x = Group, y = mean)) +
  geom_boxplot() +
  stat_summary(fun = "mean", geom = "point", shape=23, size=4, fill = "white") +
  scale_y_continuous(name = "mean(weight)")

# New plot with filtered outliers
RATSLs1 <- RATSLs %>%
  filter(mean<550 & mean>250 & mean!=495.2)

ggplot(RATSLs1, aes(x = Group, y = mean)) +
  geom_boxplot() +
  stat_summary(fun = "mean", geom = "point", shape=23, size=4, fill = "white") +
  scale_y_continuous(name = "mean(weight)")

```

```{r}
# Looking for a statistical difference between the different goups of rats diets

# Add the baseline from the original data as a new variable to the summary data
RATSLs2 <- RATSLs %>%
  mutate(baseline = RATS$WD1)

# Get rid of outliers again
RATSLs2 <- RATSLs2 %>%
  filter(mean<550 & mean>250 & mean!=495.2)

# Fit the linear model with the mean as the response 
Ratfit <- lm(mean ~ baseline + Group, data = RATSLs2)

# Compute the analysis of variance table for the fitted model with anova()
anova(Ratfit)


```

PART 2: BPRS DATASET:

Here we will work on longitudinal clinical trial data from Davis 2002, involving 40 men under suspicion of schizophrenia. The study spanned 8 weeks and utilized the Brief Psychiatric Rating Scale (BPRS) to assess 18 symptoms, including hostility, suspiciousness, and hallucinations. Ratings ranged from 1 (not present) to a maximum of 7 (severe)

```{r}
library(dplyr)
library(tidyr)
library(ggplot2)

# First let's get this data:
rm(list=ls()) # clear variables
BPRS <- read.table("https://raw.githubusercontent.com/KimmoVehkalahti/MABS/master/Examples/data/BPRS.txt", sep =" ", header = T)

# Let's have a look at this dataset BPRS
names(BPRS)
str(BPRS)

dim(BPRS) #This dataset comes with 40 observations and 11 variables.

# Print out summaries of the variables
summary(BPRS)

# We will start with turning this data into a dataframe type:
BPRSL <- as.data.frame(BPRS)

# We will turn this dataframe into a longitudinal data and extract the week number
BPRSL <- pivot_longer(BPRSL, cols = -c(treatment, subject), names_to = "weeks", values_to = "bprs") %>% arrange(weeks)
BPRSL$week <- as.integer(substr(BPRSL$weeks, 5, 5))

# "Treatment" and the "subject" are categorical so will be turned into factors:
BPRSL$treatment <- as.factor(BPRSL$treatment)
BPRSL$subject <- as.factor(BPRSL$subject)

# Have a look at the this data now:

str(BPRSL)
summary(BPRSL)

# Visualize the data using ggplot
ggplot(BPRSL, aes(x = week, y = bprs, linetype = subject)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times = 4)) +
  facet_grid(. ~ treatment, labeller = label_both) +
  theme(legend.position = "none") + 
  scale_y_continuous(limits = c(min(BPRSL$bprs), max(BPRSL$bprs)))

# Plot the RATSL data

library(ggplot2)
ggplot(BPRSL, aes(x = week, y = bprs, group = interaction(treatment, subject))) +
  geom_line(aes(linetype = treatment))+
  scale_x_continuous(name = "week")+
  scale_y_continuous(name = "bprs")+
  theme(legend.position = "top")
```

```{r}
library(lme4)

# create a regression model BPRS
BPRS_ref1 <- lm(bprs ~ week + treatment, data = BPRSL)

# print out a summary of the model
summary(BPRS_ref1)

BPRS_ref2 <- lmer(bprs ~ week + treatment + (1 | subject), data = BPRSL, REML = FALSE)

# Print the summary of the model
summary(BPRS_ref2)

BPRS_ref3 <- lmer(bprs ~ week + treatment + (week | subject), data = BPRSL, REML = FALSE)

summary(BPRS_ref3)

# perform an ANOVA test on the models tested
anova(BPRS_ref2, BPRS_ref1)

anova(BPRS_ref3,BPRS_ref2)

```

```{r}

library(dplyr)
library(tidyr)
library(ggplot2)

# Now we look at the plots of these models:
Fitted <- fitted(BPRS_ref3)

# Add the fitted model as a column
BPRSL$Fitted <- Fitted

# Create the fitted model plot of BPRSL data
ggplot(BPRSL, aes(x = week, y = Fitted, group = interaction(treatment, subject))) +
  geom_line(aes(linetype = treatment, colour=treatment)) +
  scale_x_continuous(name = "Time (weeks)", breaks = seq(0, 60, 20)) +
  scale_y_continuous(name = "Fitted BPRS") +
  theme(legend.position = "top")  +
  ggtitle("Fitted")

```
 
 Can clearly see that both treatments are effectively reducing the symptoms.